Concurrent Programming with CXL

### 2020-06-07 book/paper.tex
Added challenge
Updating terminology

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \texttt{count} variable is being updated, no other process is
trying to do the same thing.  We call this a \emph{critical section}: a
set of instructions where only one process is allowed to execute at a
time.

@@

However, a process trying to obtain a read or write lock in
Figure~ref{fig:rwbusy} has to actually obtain a spinlock just to check
if the read or write lock is available.  So even when the read or write
lock is not available, the process ends up updating the spinlock and
therefore the shared memory.

@@

CXL does not complain about Figure~\ref{fig:rwbusy}, but, using
the right test, CXL can be used to check for busy-waiting.
Figure~\ref{fig:rwblock} presents such a test.
Note that the initialization code obtains a read lock, preventing
all readers and writers from entering the critical section.
In that case, those processes should ideally be blocked and CXL
can test for that by running it with the \texttt{-b} flag.
This flag tells CXL to check
that all states are non-terminating and have all processes blocked.

@@

It's not easy.  It would be easy to come up with a solution that uses a single lock
and busy waiting, but we would like to avoid busy waiting.
Introduced by Dijkstra,
a \emph{semaphore} is a synchronization primitive that fits the bill.
A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{procure} or simply \texttt{P} operation.  The
\texttt{procure} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{vacate} or {V} operation.

@@

One can think of a semaphore as a generalization of a lock.  In fact, a
lock can be implemented by a semaphore initialized to 1.  Then \texttt{P}
acquires the lock, and \texttt{V} releases the lock.

@@

To add an item to the bounded buffer (\texttt{produce(item)}), the producing
process first has to wait until there is room in the bounded buffer.
To this end, it invokes \texttt{P(n\_empty)}, which waits until
\texttt{n\_empty > 0} and atomically decrements \texttt{n\_empty} once this
is the case.  Next, it adds the item to the next position in the buffer.
Since there may be more than one empty slot, multiple producers may attempt
to do the same thing concurrently.  Hence the use of the \texttt{l\_in}
semaphore used as a lock.  Finally, once the item is added, the
\texttt{n\_full} semaphore is incremented to signal to consumers that
an item is available.

@@

@@

>>>>

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \texttt{count} variable is being updated, no other process is
trying to do the same thing.  We call this a \emph{critical section}
(aka critical region): a
set of instructions where only one process is allowed to execute at a
time.

@@

However, a process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to actually obtain a spinlock just to check
if the read or write lock is available.  So even when the read or write
lock is not available, the process ends up updating the spinlock and
therefore the shared memory.

@@

CXL does not complain about Figure~\ref{fig:rwbusy}, but, using
the right test, CXL can be used to check for busy-waiting.
Figure~\ref{fig:rwblock} presents such a test.
Note that the initialization code obtains a read lock, preventing
all readers and writers from entering the critical section.
In that case, those processes should ideally be blocked and CXL
can test for that by running it with the \texttt{-b} flag.
This flag tells CXL to check
that all states are non-terminating and lead to a state in which
all processes are blocked.

@@

It's not easy.  It would be easy to come up with a solution that uses a single lock
and busy waiting, but we would like to avoid busy waiting.
Introduced by Dijkstra,
a \emph{semaphore} is a synchronization primitive that fits the bill.
A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{P} operation.  You can think of \texttt{P} standing
for \emph{Procure}, as in procuring the resource associated with the semaphore.
The \texttt{P} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{V} operation.  You can think of \texttt{V} standing for \emph{vacate},
as in vacating the resource.

@@

One can think of a semaphore as a generalization of a lock.  In fact, a
lock can be implemented by a semaphore initialized to 1.  Then \texttt{P}
procures the lock, and \texttt{V} vacates the lock.

@@

To add an item to the bounded buffer (\texttt{produce(item)}), the producing
process first has to wait until there is room in the bounded buffer.
To this end, it invokes \texttt{P(n\_empty)}, which waits until
\texttt{n\_empty > 0} and atomically decrements \texttt{n\_empty} once this
is the case.  In other words, the producer procures an empty slot.
Next, it adds the item to the next position in the buffer.
Since there may be more than one empty slot, multiple producers may attempt
to do the same thing concurrently.  Hence the use of the \texttt{l\_in}
semaphore used as a lock.  Finally, once the item is added, the process
increments the \texttt{n\_full} semaphore to signal to consumers that
an item is available.

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWbinsem.cxl}
\end{code}
\caption{Reader/Writer Lock implementation using Split Binary Semaphores.}
\label{fig:RWsplitsema}
\end{figure}

Reader/writer locks and bounded buffers are both examples of what are
sometimes called \emph{conditional critical sections} (CCSs).
All critical sections are in fact conditional.  Even the basic one
has the condition that a process can only enter if no other process
is in the critical section.  Reader/writer locks refined that, loosening
the restriction for improved performance.
In the case of bounded buffers, there were two CCSs.  First, a CCS
to add an entry to the bounded buffer that can only be entered by one
process and only if there is room in the bounded buffer (\texttt{produce}).
Second, a CCS that can only be entered by one process and
only if there is an item in the bounded buffer (\texttt{consume}).

CCSs are easy to implement using busy waiting:
\begin{code}
\begin{verbatim}
let blocked = True:
    while blocked:
        call lock();
        if application-condition-holds:
            blocked = False;
        else:
            call unlock():
@cr: application-code;
call unlock():
\end{verbatim}
\end{code}

We have seen solutions to the reader/writer problem and the producer/consumer
problem that do not busy-wait, but they are quite different.  It would be nice
to have a general technique for all CCS problems.
The ``Split Binary Semaphore'' (SBS) is such a technique, originally proposed by
Tony Hoare and popularized by Edsger Dijkstra.
SBS is an example of what one could call a ``baton-passing'' technique.
A process needs the baton to access shared variables.
When a process has the baton and does not need it any more,
it first checks to see if there are any processes that are waiting for the
baton and can make progress.
If so, it passes the baton directly to one of those processes.
If not, it simply gives up the baton.

We will illustrate the technique using the reader/writer problem.
Figure~\ref{fig:RWsplitsema} shows the
\texttt{rlock\_acquire}, \texttt{rlock\_release},
\texttt{wlock\_acquire}, and \texttt{wlock\_release} methods implemented using
SBS.

Step 1 is to enumerate all waiting conditions.  In the case of the reader/writer
problem, there are two: a process that wants to read may have to wait for a
writer to leave the critical regiion, while a process that wants to write may
have to wait until all readers have left the critical section.  If there are $N$
waiting conditions, the SBS technique will use $N+1$ binary semaphores: one for
each waiting condition and an extra one that processes use when they first try
to enter the critical section and do not yet know if they need to wait.
We shall call this the \texttt{mutex}.

An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
must always be 0 or 1, and it should be 0 when a process is accessing the shared
variables.
As it were, these semaphores taken together are emulating a single binary semaphore.
Initially, \texttt{mutex} is~1 and the other $N$ semaphores
are all~0.  To maintain the invariant, each process alternates \texttt{P} and
\texttt{V} operations, starting with \texttt{P~\&(mutex)} when it tries to
enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
(yet to be determined).

Another important part of the SBS technique is to keep careful track of the
state.  In the case of the reader/writer lock, the state consists of the
following shared variables:
\begin{itemize}
\item \texttt{nreadersEntered}: the number of readers in the critical section;
\item \texttt{nwritersEntered}: the number of writers in the critical section;
\item \texttt{nreadersWaiting}: the number of readers waiting to enter;
\item \texttt{nwritersWaiting}: the number of writers waiting to enter.
\end{itemize}
Each of the
\texttt{rlock\_acquire}, \texttt{rlock\_release},
\texttt{wlock\_acquire}, and \texttt{wlock\_release} methods must maintain
this state.
To wit, check out \texttt{rlock\_acquire}.  It first checks to see if there
are any writers.  If so, it increments \texttt{nreadersWaiting}.  If not,
it increments \texttt{nreadersEntered}.

Either way, it vacates one of the semaphores next.
\texttt{V\_one()} is the baton passing function that selects which of the
binary semaphores to vacate.
Method \texttt{V\_one()} first checks to see if there are any readers or
writers waiting.  If there are readers waiting and there are no writers
in the critical section, it vacates the semaphore associated with
readers waiting.  If there are writers waiing and there are no readers
nor writers in the critical section, it vacates the semaphore associated
with writers waiting.  If both conditions hold, it could vacates either
one---it would not matter.  However, it can only vacate one of the semaphores.
If none of the conditions hold, \texttt{V\_one()} vacates \texttt{mutex}.

Let us return now to \texttt{rlock\_acquire}, in the case that the process
found there is a writer in the critical section, it vacates
\texttt{mutex} and procures \texttt{rsema},
blocking on the semaphore associated
with waiting readers.  When later some other process passes the baton to
it by incrementing \texttt{rsema}), it updates the state by decrementing
\texttt{nreadersWaiting} and incrementing \texttt{nreadersEntered}.  It
then invokes \texttt{V\_one()} one more time.

As an example, consider the case where there is a writer in the critical
section and there are two readers waiting.  Let us see what happens when
the writer calls \texttt{wlock\_release}:
\begin{enumerate}
\item After procuring \texttt{mutex},
the sum of the three semaphores is 0.  The writer then decrements
\texttt{nwritersEntered} and calls \texttt{V\_one()}.
\item \texttt{V\_one()} finds that there are no writers in the critical section
and there are two readers waiting.  It therefore vacates \texttt{rsema}.
The sum of the semaphores is now 1, because \texttt{rsema} is 1.  This
guarantees that only a waiting reader, procuring \texttt{rsema}, can enter
the critical section.
\item When it does, the reader decrements \texttt{nreadersWaiting}
from 2 to 1, and increments \texttt{nReadersEntered} from 0 to 1.
The reader finally calls \texttt{V\_one()}.
\item Again, \texttt{V\_one()} finds that there are no writers and
that there are readers waiting, so again it vacates \texttt{rsema},
passing the baton to the one remaining waiting reader.
\item The remaining reader decrements \texttt{nreadersWaiting} from 1 to 0 and
increments \texttt{nReadersEntered} from 1 to 2.
\item Finally, the remaining reader  calls \texttt{V\_one()}.
\texttt{V\_one()} does not find any processes waiting,
and so it vacates \texttt{mutex}.
\end{enumerate}

\begin{quote}
\begin{itemize}
\item Implement the producer/consumer problem with SBS.
\end{itemize}
\end{quote}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../RWfair.cxl}
\end{code}
\caption{Reader/Writer Lock implementation addressing fairness.}
\label{fig:RWfair}
\end{figure}
